Clinical Trial Data Analysis

### 2016-12-09 results/results.Rmd
Untrack old data analysis files; update data analysis to use cleaned dataset; add cache files to gitignore

---
title: "RB Data"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

ga <- read.csv("GA1102.csv", header = TRUE, stringsAsFactors = FALSE)
```

@@

### Missing data
There's a bunch of missing data. 
First, we notice there are 14 subjects who were only observed during the baseline period.

```{r baseline_only, results = "asis"}
numdays <- table(ga$SUBJID)
baseline_only <- names(numdays)[numdays < 14]
unique_numdays <- as.data.frame(table(numdays))
names(unique_numdays) <- NULL
print(xtable(unique_numdays,
      caption = "Number of subjects observed for 7 or 14 days"),
      include.rownames = FALSE, comment = FALSE)

ga$missing <- ifelse(ga$SUBJID %in% baseline_only, 1, 0)
```

Next, we check how many individuals were assigned to each treatment group.
52 subjects have treatment values of `.`.
They appear to never have been surveyed during week 2: although there are records in the dataset, all values are 0.
These 52 include the fourteen who were only observed during week 1.
We remove all these missing subjects.
We are left with a total of 66 subjects in the placebo group and 70 subjects in the active treatment group.

```{r baseline_only2, results = "asis"}
num_treated <- table(ga$tr)
baseline_only2 <- as.data.frame(num_treated/14)
colnames(baseline_only2) <- c("Treatment", "Number of subjects")
print(xtable(baseline_only2, digits = 0,
             caption = "Number of subjects in each treatment group"), 
      include.rownames = FALSE, comment = FALSE)

ga$missing <- ga$missing + ifelse(ga$tr == ".", 1, 0)
```

## Exploratory Data Analysis

We notice several things patterns from Figure \ref{fig:visualize_missing}, showing missing data patterns.
First, some of the sites have many more subjects than others: these are sites 1, 4, and 12.
Second, some of the sites have a large proportion of missing data, such as sites 3, 4, 5, and 14.
Finally, we notice that sites 3, 6, and 14 have very few subjects, so we will have low power at these sites.
For example, after removing the missing values, site 14 only has one treated subject and two placebos, so the site will only have 3 unique permutation statistics.

```{r visualize_missing, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
ga %>%
  mutate(subjid_by_site = SUBJID %% 100) %>%
  ggplot(aes(x = subjid_by_site, y = factor(SITEID))) + 
  geom_tile(aes(fill = tr)) + 
  xlab("Subject ID Number") +
  ylab("Site ID") + 
  scale_fill_discrete(name="Treatment",
                      breaks=c(".", "1", "2"),
                      labels=c("Not assigned", "Placebo", "Drug")) +
  ggtitle("Treatment Assignment Patterns")
```

```{r remove_missing, include = FALSE}
ga <- ga %>% filter(tr != ".") %>% mutate(tr = as.numeric(tr)-1)
```

@@

data_by_subjid_visitnum <- ga %>% group_by(SUBJID, VISITNUM)

reshape_ga_continuous <- function(variable, data = data_by_subjid_visitnum){
  data <- data %>% mutate_(myvariable = variable)
  summarized <- data %>% summarise("variable" = mean(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
  cleaned <- ungroup(cleaned) %>% 
    mutate(difference = Outcome - Baseline)
  return(cleaned)
}

```
Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between treatment and placebo groups.
Based on these distributions, we would expect to detect a difference in outcomes between the drug and placebo groups for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.

```{r hist_continuous_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of continuous outcomes"}
continuous_vars <- c("daily_heart", "daily_regurg", "daily_dysp", 
                    "daily_hrdq",
                    "heart_freq", "regurg_freq", "dysp_freq")
plot_distrs <- lapply(continuous_vars, function(variable){
  p <- reshape_ga_continuous(variable) %>% mutate(tr = factor(tr)) %>%
        ggplot(aes(Outcome)) +
        geom_density(alpha = 0.6, aes(fill = tr))+ 
#        geom_histogram(binwidth = 0.5, position = "dodge", aes(fill = tr)) +
        scale_fill_discrete(name="Experimental\nCondition",
                       breaks=c("0", "1"),
                       labels=c("Placebo", "Drug")) + 
        xlab(variable)
  return(p)
})

@@

### Ordinal variables

Several variables are measured on an ordinal scale but treated as continuous values.
For instance, heartburn frequency is measured on a 7-point scale, with levels 0, 1, 2, 3, 4-5, 6-10, and more than 10.
Values in the 4-5 range are coded as 4.5 and in the 6-10 range are coded as 8.
We do not necessarily believe that these are accurate representations of the data.
We recode the ordinal scales as factors on a scale from 1 to 7.

```{r recode}
recode_ordinal <- function(variable){
  levels <- 
  variable <- as.factor(variable)
  levels(variable) <- seq_along(unique(variable))
  return(variable)
}

ga <- ga %>% mutate(heart_sev_f = recode_ordinal(heart_sev),
                    regurg_sev_f = recode_ordinal(regurg_sev),
                    dysp_sev_f = recode_ordinal(dysp_sev),
                    heart_freq_f = recode_ordinal(heart_freq),
                    regurg_freq_f = recode_ordinal(regurg_freq),
                    dysp_freq_f = recode_ordinal(dysp_freq)
                    )
```
We must put the data in a usable form so there is only one measurement for each individual.
For each person, we take the median response from week 2 and subtract it from their median response from week 1.

```{r data_structure_ordinal}

data_by_subjid_day <- ga %>% 
  mutate(dayofweek = rep(1:7, length.out = nrow(ga))) %>%
  group_by(SUBJID, dayofweek)

# The reshape function below reduces the 14 days into a single measure with the following steps:
# 1) Take the difference in measurements between the first day of treatment week and the first day of baseline week, the second day of treatment week and the second day of baseline week, and so on.
# 2) Take the median of these seven numbers.
# We have decided not to use this method; it assumes that day j in weeks 1 and 2 are paired, which may not be right.

# reshape_ga_ordinal <- function(variable, data = data_by_subjid_day){
#   data <- data %>% mutate_(myvariable = variable)
#   day_diff <- data %>% 
#     summarise("variable" = diff(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
#   summarized <- day_diff %>%
#     summarise("medoverdays" = median(variable), "tr" = unique(tr), "SITEID" = unique(SITEID))
#   summarized <- ungroup(summarized)
#   return(summarized)
# }
reshape_ga_ordinal <- function(variable, data = data_by_subjid_visitnum){
  data <- data %>% mutate_(myvariable = variable)
  summarized <- data %>% summarise("variable" = median(as.numeric(myvariable)), "tr" = unique(tr), "SITEID" = unique(SITEID))
  cleaned <- dcast(summarized, SUBJID + tr + SITEID ~ VISITNUM, value.var = "variable")
  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
  cleaned <- ungroup(cleaned) %>% 
    mutate(difference = Outcome - Baseline)
  return(cleaned)
}
```
Figure \ref{fig:hist_ordinal_vars} displays the distribution of these newly-coded ordinal variables in the treatment and control groups.
There are certainly differences in distributions, but it is not clear how different they are.

```{r hist_ordinal_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of ordinal outcomes"}
ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f") 
plot_distrs <- lapply(ordinal_vars, function(variable){
  p <- reshape_ga_ordinal(variable) %>% 
        mutate(tr = factor(tr)) %>%
        ggplot(aes(difference, ..density..)) +
        geom_bar(aes(y = (..count..)/sum(..count..), fill = tr), position = "dodge") + 
        scale_y_continuous(labels = scales::percent) +
        scale_fill_discrete(name="Experimental\nCondition",
                       breaks=c("0", "1"),
                       labels=c("Placebo", "Drug")) +
        xlab(variable)
  return(p)
})

plot_distrs <- lapply(plot_distrs, function(x) x + theme(legend.position="none"))
plot_distrs[[length(plot_distrs)+1]] <- legend
do.call(grid.arrange, c(plot_distrs, nrow = 3))
```

# Analysis - Primary endpoint
Let's first restrict attention to the primary endpoint, `daily_hrdq`.
We must transform the data to contain the variables of interest: the baseline score (averaged over 7 days of no treatment), the outcome score (averaged over 7 days of treatment), an indicator for which treatment was received (`1` for the active drug, `0` for placebo), the site ID, and subject ID.

Throughout, we compare two models: 
the first model considers the final outcome measure, adjusting for site ID and perhaps also the baseline score.
The second model considers the difference between outcome and baseline score as its dependent variable, adjusting for site ID.

We compare three testing methods: 
the usual parametric ANOVA,
a permutation test considering the difference in means as test statistic, stratifying by site,
and a permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.

```{r hrdq, results = "asis"}
daily_hrdq <- reshape_ga_continuous("daily_hrdq")
summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == 1]),
                            summary(daily_hrdq$difference[daily_hrdq$tr == 0]))
colnames(summary_stats_hrdq) <- c("Drug", "Placebo")
print(xtable(summary_stats_hrdq,
             caption = "Summary statistics for change in primary endpoint (daily hrdq) for drug and placebo groups."),
      include.rownames = TRUE)
```

## Parametric ANOVA
First, we do a standard parametric ANOVA using the two models.
Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect.
Under model 2, the significance goes away.

```{r hrdq_parametric, results = "asis"}
# method 0: parametric ANOVA
lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm1))), include.rownames = TRUE)
daily_hrdq$lm1_resid <- residuals(lm1)

@@

Suppose each individual, $i = 1, \dots, N$ has two potential outcomes $(Y_i(1), Y_i(0))$ that indicate their response to the active drug and to the placebo, respectively.
We randomly assign treatment to individuals; treatment determines which of $Y_i(1)$ and $Y_i(0)$ we observe.
We are unable to observe both.
Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
This is referred to as the ``sharp null'' hypothesis:

$$H_0: Y_i(1) = Y_i(0), i = 1, \dots, N.$$

Then, whether individual $i$ received the drug or placebo amounts to an arbitrary label.
Once we observe their response under treatment $T_i$, we know what it would have been under treatment $1-T_i$, for $T_i \in \{0,1\}$; namely, it would be the same.

Treatment was completely randomized at each site, independently across sites.
Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received the drug and who received the placebo at each site.
Any assignment of treatments which preserves the number of treated and controls at each site is valid and has equal probability of occurring.
Therefore, by using this principle of equal probabilities and by imputing the unobserved potential outcomes assuming that the null hypothesis is true, we may obtain the permutation distribution of any statistic under the null hypothesis.

We compare two test statistics: the usual difference in means between treatment and placebo groups, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
The first statistic is more comparable to what one would obtain from an ANOVA and it is readily interpretable.
It does not directly account for the stratification by site, so variation between sites may be hidden.
The second statistic is useful for testing the two-sided alternative hypothesis that treatment has some effect. 
It may be more powerful than the simple difference in means if the effect of the drug varies across sites.
For instance, if the drug has a positive effect at one site but a negative effect at another site, this test statistic would be large; 
if we used the simple difference in means without accounting for sites, then the positive effect and negative effect may cancel each other out and appear as no effect.

@@

```{r hrdq_unadjusted_permutation, fig.height = 3, fig.cap = "Permutation distribution of the difference of means for daily_hrdq"}
# method 1 : do permutation of differences
observed_diff_means1 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$Outcome) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$Outcome)
diff_means_distr1 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$Outcome, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue1 <- t2p(observed_diff_means1, diff_means_distr1, alternative = "two-sided")

observed_diff_means2 <- mean(daily_hrdq[daily_hrdq$tr == 1, ]$difference) - mean(daily_hrdq[daily_hrdq$tr == 0, ]$difference)
diff_means_distr2 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$difference, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

@@

## Covariate-adjusted permutation test
We would like to test for a difference in outcomes between the active drug and placebo group, but control for other covariates.
In particular, when the outcome is the average response during the second week of follow-up, we would like to control for the average response during the first week of follow-up and the location.
When the outcome is the difference in responses between the second and first weeks, we would just like to control for location.
We will use the approximate permutation test derived by Freedman and Lane (1983) to do so.

Let $Y$ denote the response, $Z$ denote the treatment indicator, and $X$ denote a covariate which may be correlated with $Y$.
We may write the following equation:

@@

observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
lm1_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
})

@@

observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
lm2_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm2_perm <- lm(difference ~ tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["tr_perm", "t value"]
})

@@

lm1_no_tr <- lm(Outcome~Baseline + factor(SITEID), data = daily_hrdq)
lm1_resid <- residuals(lm1_no_tr)
lm1_yhat <- fitted(lm1_no_tr)
observed_t1 <- summary(lm1)[["coefficients"]]["tr", "t value"]
lm1_t_distr <-  replicate(1e4, {
  lm1_resid_perm <- permute_within_groups(lm1_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm1_yhat + lm1_resid_perm
  lm1_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["tr", "t value"]
})

@@

lm2_no_tr <- lm(difference ~ factor(SITEID), data = daily_hrdq)
lm2_resid <- residuals(lm2_no_tr)
lm2_yhat <- fitted(lm2_no_tr)
observed_t2 <- summary(lm2)[["coefficients"]]["tr", "t value"]
lm2_t_distr <-  replicate(1e4, {
  lm2_resid_perm <- permute_within_groups(lm2_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm2_yhat + lm2_resid_perm
  lm2_perm <- lm(response_fl ~ tr + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["tr", "t value"]
})

@@

outcome_pvalues <- c(summary(aov(lm1))[[1]]["tr", "Pr(>F)"],
                     diff_means_pvalue1,
                     diff_means_bystrata_pvalue1,
                     lm_pvalue_1,
                     fl_pvalue_1)
differenced_pvalues <- c(summary(aov(lm2))[[1]]["tr", "Pr(>F)"],
                     diff_means_pvalue2,
                     diff_means_bystrata_pvalue2,
                     lm_pvalue_2,
                     fl_pvalue_2)
pvalues_table <- cbind(differenced_pvalues, outcome_pvalues)

@@

for(col in continuous_vars){
  i <- i+1
  tmpdata <- reshape_ga_continuous(col)
  lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
  pvalues_table_contin[i, 1] <- summary(aov(lm1))[[1]]["tr", "Pr(>F)"]

  observed_diff_means <- mean(tmpdata[tmpdata$tr == 1, ]$Outcome) - mean(tmpdata[tmpdata$tr == 0, ]$Outcome)
  diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e3)
  pvalues_table_contin[i, 2] <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
  lm1_t_distr <-  replicate(1e3, {
    tmpdata$tr_perm <- permute_within_groups(tmpdata$tr, tmpdata$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = tmpdata)
  summary(lm1_perm)[["coefficients"]]["tr_perm", "t value"]
  })
  pvalues_table_contin[i, 3] <- t2p(observed_t, lm1_t_distr, alternative = "two-sided")

  lm_no_tr <- lm(Outcome ~ Baseline + factor(SITEID), data = tmpdata)
  lm_resid <- residuals(lm_no_tr)
  lm_yhat <- fitted(lm_no_tr)
  observed_t <- summary(lm1)[["coefficients"]]["tr", "t value"]
  lm_t_distr <-  replicate(1e3, {
      lm_resid_perm <- permute_within_groups(lm_resid, tmpdata$SITEID)
      tmpdata$response_fl <-  lm_yhat + lm_resid_perm
      lm_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = tmpdata)
      summary(lm_perm)[["coefficients"]]["tr", "t value"]
})

@@

# Analysis - Ordinal variables

For ordinal variables, it doesn't make sense to do ANOVA.
Instead, we do a multi-aspect analysis to test whether there is any difference in the outcome distributions between the drug and placebo groups.
For this, we use the same stratified permutation test approach as above.
However, we use three different test statistics and combine them into a single, global test using the nonparametric combination (NPC) method.
The test statistics are the difference in mean outcomes between the drug and placebo group, the Anderson-Darling statistic, and the Kolmogorov-Smirnov statistic.
We use the Tippett combining function.

The outcome measures used are the difference in median measures during the treatment week minus the median measures during the baseline week.
Recall from Figure \ref{fig:hist_ordinal_vars} that it is not clear whether the distributions are different.
Table \ref{ordinal_results_print} shows the p-values for the three partial tests plus the global p-value.
None is significant.
It appears that treating the variables as ordinal results in a loss of information and, consequently, power.

```{r ordinal_tests}
# Compute difference in mean value (median outcome) between treatment and placebo
mean_median <- function(outcome, treatment) {
  tr_levels <- sort(unique(treatment))
  mean(outcome[treatment == tr_levels[2]], na.rm = TRUE) - 
    mean(outcome[treatment == tr_levels[1]], na.rm = TRUE)
  }

# Anderson-Darling
ad<-function(outcome, treatment){    
  n <- length(treatment)   
  W <- table(outcome, treatment)
  ecdf <- apply(W, 2, cumsum)
  ecdf_comb <- apply(ecdf, 1, sum)

  ratio <- ecdf[,2]/(ecdf_comb*(n-ecdf_comb))
  j <- length(ratio)
  T_ad <- sum(ratio[seq_len(j-1)])
  return(T_ad)
}

# Kolmogorov-Smirnov
ks <- function(outcome, treatment){
  tr_levels <- sort(unique(treatment))
  # Assume that larger treatment level is active treatment, smaller other is control
  g1 <- outcome[treatment == tr_levels[2]]
  g0 <- outcome[treatment == tr_levels[1]]
  suppressWarnings(ks.test(g1, g0)$statistic)
}

multiaspect_test <- function(outcome, treatment, strata, reps = 1000){
  obs_ts <- c(mean_median(outcome, treatment),
              ad(outcome, treatment),
              ks(outcome, treatment))
  names(obs_ts) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
  perm_distr <- t(replicate(reps, {
            perm_treatment <- permute_within_groups(treatment, strata)
            c(mean_median(outcome, perm_treatment),
              ad(outcome, perm_treatment),
              ks(outcome, perm_treatment))
            }))

  partial_pvalues <- sapply(seq_along(obs_ts), function(i){
    t2p(obs_ts[i], perm_distr[, i], "two-sided")
  })
  names(partial_pvalues) <- c("Difference in Means", "Anderson-Darling", "Kolmogorov-Smirnov")
  global_pvalue <- npc(obs_ts, perm_distr, alternatives="two-sided", combine = "tippett")

  return(list(
    "Partial test statistics" = obs_ts,
    "Partial p-values"        = partial_pvalues,
    "Global p-value"          = global_pvalue
  ))
}
```

```{r ordinal_results_make}
set.seed(609726531) # Generated from Random.org Timestamp: 2016-11-09 15:38:50 UTC
ordinal_vars <- c("heart_sev_f", "regurg_sev_f", "dysp_sev_f",
                    "heart_freq_f", "regurg_freq_f", "dysp_freq_f")

pvalues_table_ord <- as.data.frame(matrix(NA, nrow = length(ordinal_vars), ncol = 4))
i <- 0

>>>>

---
title: "Clinical Trial Data Analysis"
author: "Kellie Ottoboni"
date: "`r Sys.Date()`"
output: pdf_document
---

@@

clinical <- read.csv("../data/clinical_cleaned.csv", header = TRUE, stringsAsFactors = FALSE)
```

@@

We preprocessed the data from its original format in the file "data/clinical_cleaned.R" so that it would be sufficiently anonymized.
We took the following steps:

* Initially, there were 188 subjects. 52 of these were never assigned to a treatment group and were only observed at baseline. This ensures that censoring was independent of treatment, so removing them from the analysis does not introduce any selection bias. We are left with 66 subjects in group A and 70 subjects in group B.
* We renamed treatments, subject IDs, and site IDs to remove identifying information.
* Each individual was observed for a total of 14 days, 7 during the baseline week and 7 after receiving treatment.  We averaged the 7 measurements taken during each of these time periods, leaving us with 2 measurements per individual.

# Exploratory Data Analysis

We notice several things patterns from Figure \ref{fig:count_by_site}.
Some of the sites have many more subjects than others: these are sites 1, 3, and 6, and 7.
Sites 2, 5, and 8 have very few subjects, so we will have low power at these sites.
For example, after removing the missing values, site 8 only has one person in group B and two in group A, so the site will only have 3 unique permutation statistics.

```{r count_by_site, fig.align="center", fig.cap = "Number of subjects in each treatment group, by site ID."}
clinical %>% 
  filter(VISITNUM == 1) %>% 
  ggplot(aes(x = factor(SITEID))) + 
  geom_bar(aes(fill = tr), position = "dodge") +
  xlab("Site ID") +
  ylab("Count") +
  ggtitle("Number of Individuals per Treatment") +
  labs(fill = "Treatment") 
```

@@

data_by_subjid_visitnum <- clinical %>% group_by(SUBJID, VISITNUM)

reshape_data <- function(variable, data = data_by_subjid_visitnum){
  data <- data %>% mutate_(myvariable = variable)
  cleaned <- dcast(data, SUBJID + tr + SITEID ~ VISITNUM, value.var = "myvariable")
  colnames(cleaned) <- c("SUBJID", "tr", "SITEID", "Baseline", "Outcome")
  cleaned <- ungroup(cleaned) %>% 
    mutate(difference = Outcome - Baseline)
  return(cleaned)
}

```
Figure \ref{fig:hist_continuous_vars} shows the distribution of the final outcome measures between the two treatment groups.
Based on these distributions, we would expect to detect a difference in outcomes between the treatments for the `daily_heart`, `daily_hrdq`, and `heart_freq` measures.

```{r hist_continuous_vars, fig.height=6, fig.width=12, fig.cap = "Distribution of continuous outcomes"}
continuous_vars <- c("daily_heart", "daily_regurg", "daily_dysp", 
                    "daily_hrdq",
                    "heart_freq", "regurg_freq", "dysp_freq")
plot_distrs <- lapply(continuous_vars, function(variable){
  p <- reshape_data(variable) %>% mutate(tr = factor(tr)) %>%
        ggplot(aes(Outcome)) +
        geom_density(alpha = 0.6, aes(fill = tr))+ 
#        geom_histogram(binwidth = 0.5, position = "dodge", aes(fill = tr)) +
#        scale_fill_discrete(name="Experimental\nCondition",
#                       breaks=c("0", "1"),
#                       labels=c("A", "B")) + 
        labs(x = variable, fill = "Treatment")
  return(p)
})

@@

# Analysis - Primary endpoint
To illustrate the different methods, let's first restrict attention to the primary endpoint, `daily_hrdq`.
Throughout, we compare two models: 
the first model considers the final outcome measure, adjusting for site ID and perhaps also the baseline score.
The second model considers the difference between outcome and baseline score as its dependent variable, adjusting for site ID.

We compare four testing methods: 
the usual parametric ANOVA,
a permutation test considering the difference in means as test statistic, stratifying by site,
the stratified permutation test, using the $t$-statistic from the linear regression instead,
and a permutation test in the style of Freedman and Lane (1983), using a linear regression to adjust for covariates.

```{r hrdq, results = "asis"}
daily_hrdq <- reshape_data("daily_hrdq")
summary_stats_hrdq <- cbind(summary(daily_hrdq$difference[daily_hrdq$tr == "B"]),
                            summary(daily_hrdq$difference[daily_hrdq$tr == "A"]))
colnames(summary_stats_hrdq) <- c("B", "A")
print(xtable(summary_stats_hrdq,
             caption = "Summary statistics for change in primary endpoint (daily hrdq) for the two treatment groups."),
      include.rownames = TRUE)
```

## Parametric ANOVA
First, we do a standard parametric ANOVA using the two models.
Under model 1, using the outcome as response and adjusting for the baseline value, treatment has a significant effect.
Under model 2, the significance goes away.
(Note that treatment is still coded as A and B, which are characters. The `lm` function is smart enough to know that it should be coded as a factor, where we compare the effect of B relative to A. In general, be very careful when you're doing this!)
```{r hrdq_parametric, results = "asis"}
# method 0: parametric ANOVA
lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
print(xtable(summary(aov(lm1))), include.rownames = TRUE)
daily_hrdq$lm1_resid <- residuals(lm1)

@@

Suppose each individual, $i = 1, \dots, N$ has two potential outcomes $(Y_i(A), Y_i(B))$ that indicate their response to treatments A and B, respectively.
We randomly assign treatment to individuals; treatment determines which of $Y_i(A)$ and $Y_i(B)$ we observe.
We are unable to observe both.
Suppose we wish to test the null hypothesis that individual by individual, treatment has no effect.
This is referred to as the ``sharp null'' hypothesis:

$$H_0: Y_i(A) = Y_i(B), i = 1, \dots, N.$$

Then, whether individual $i$ received A or B amounts to an arbitrary label.
Once we observe their response under one treatment, we know what it would have been under the other; namely, it would be the same.

Treatment was completely randomized at each site, independently across sites.
Assuming that drop-out from the study was independent of treatment assignment (which is true if treatment was blinded and assigned at random), then we may condition on the number of individuals who received A and B at each site.
Any assignment of treatments which preserves the number of treated and controls at each site is valid and has equal probability of occurring.
Therefore, by using this principle of equal probabilities and by imputing the unobserved potential outcomes assuming that the null hypothesis is true, we may obtain the permutation distribution of any statistic under the null hypothesis.

We compare two test statistics: the usual difference in means between group B and group A, and the difference in means within each site, aggregated by taking the sum of their absolute values over sites.
The first statistic is more comparable to what one would obtain from an ANOVA and it is readily interpretable.
It does not directly account for the stratification by site, so variation between sites may be hidden.
The second statistic is useful for testing the two-sided alternative hypothesis that treatment has some effect. 
It may be more powerful than the simple difference in means if the effect of the drug varies across sites.
For instance, if the drug has a positive effect at one site but a negative effect at another site, this test statistic would be large; 
if we used the simple difference in means without accounting for sites, then the positive effect and negative effect may cancel each other out and appear as no effect.

@@

```{r hrdq_unadjusted_permutation, fig.height = 3, fig.cap = "Permutation distribution of the difference of means for daily_hrdq"}
# method 1 : do permutation of differences
observed_diff_means1 <- mean(daily_hrdq[daily_hrdq$tr == "B", ]$Outcome) - mean(daily_hrdq[daily_hrdq$tr == "A", ]$Outcome)
diff_means_distr1 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$Outcome, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue1 <- t2p(observed_diff_means1, diff_means_distr1, alternative = "two-sided")

observed_diff_means2 <- mean(daily_hrdq[daily_hrdq$tr == "B", ]$difference) - mean(daily_hrdq[daily_hrdq$tr == "A", ]$difference)
diff_means_distr2 <- stratified_two_sample(group = daily_hrdq$tr, response = daily_hrdq$difference, stratum = daily_hrdq$SITEID, reps = 1e4)
diff_means_pvalue2 <- t2p(observed_diff_means2, diff_means_distr2, alternative = "two-sided")

@@

## Covariate-adjusted permutation test
We would like to test for a difference in outcomes between the treatment groups group, but control for other covariates.
In particular, when the outcome is the average response during the second week of follow-up, we would like to control for the average response during the first week of follow-up and the location.
When the outcome is the difference in responses between the second and first weeks, we would just like to control for location.
We will use the approximate permutation test derived by Freedman and Lane (1983) to do so.

Let $Y$ denote the response, $Z$ denote the treatment indicator ($1$ if the person gets treatment B, $0$ if they get A), and $X$ denote a covariate which may be correlated with $Y$.
We may write the following equation:

@@

observed_t1 <- summary(lm1)[["coefficients"]]["trB", "t value"]
lm1_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["tr_permB", "t value"]
})

@@

observed_t2 <- summary(lm2)[["coefficients"]]["trB", "t value"]
lm2_t_distr <-  replicate(1e4, {
  daily_hrdq$tr_perm <- permute_within_groups(daily_hrdq$tr, daily_hrdq$SITEID)
  lm2_perm <- lm(difference ~ tr_perm + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["tr_permB", "t value"]
})

@@

lm1_no_tr <- lm(Outcome~Baseline + factor(SITEID), data = daily_hrdq)
lm1_resid <- residuals(lm1_no_tr)
lm1_yhat <- fitted(lm1_no_tr)
observed_t1 <- summary(lm1)[["coefficients"]]["trB", "t value"]
lm1_t_distr <-  replicate(1e4, {
  lm1_resid_perm <- permute_within_groups(lm1_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm1_yhat + lm1_resid_perm
  lm1_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = daily_hrdq)
  summary(lm1_perm)[["coefficients"]]["trB", "t value"]
})

@@

lm2_no_tr <- lm(difference ~ factor(SITEID), data = daily_hrdq)
lm2_resid <- residuals(lm2_no_tr)
lm2_yhat <- fitted(lm2_no_tr)
observed_t2 <- summary(lm2)[["coefficients"]]["trB", "t value"]
lm2_t_distr <-  replicate(1e4, {
  lm2_resid_perm <- permute_within_groups(lm2_resid, daily_hrdq$SITEID)
  daily_hrdq$response_fl <-  lm2_yhat + lm2_resid_perm
  lm2_perm <- lm(response_fl ~ tr + factor(SITEID), data = daily_hrdq)
  summary(lm2_perm)[["coefficients"]]["trB", "t value"]
})

@@

outcome_pvalues <- c(summary(aov(lm1))[[1]]["trB", "Pr(>F)"],
                     diff_means_pvalue1,
                     diff_means_bystrata_pvalue1,
                     lm_pvalue_1,
                     fl_pvalue_1)
differenced_pvalues <- c(summary(aov(lm2))[[1]]["trB", "Pr(>F)"],
                     diff_means_pvalue2,
                     diff_means_bystrata_pvalue2,
                     lm_pvalue_2,
                     fl_pvalue_2)
pvalues_table <- cbind(differenced_pvalues, outcome_pvalues)

@@

for(col in continuous_vars){
  i <- i+1
  tmpdata <- reshape_data(col)
  lm1 <- lm(Outcome ~ Baseline + tr + factor(SITEID), data = tmpdata)
  pvalues_table_contin[i, 1] <- summary(aov(lm1))[[1]]["tr", "Pr(>F)"]

  observed_diff_means <- mean(tmpdata[tmpdata$tr == "B", ]$Outcome) - mean(tmpdata[tmpdata$tr == "A", ]$Outcome)
  diff_means_distr <- stratified_two_sample(group = tmpdata$tr, response = tmpdata$Outcome, stratum = tmpdata$SITEID, reps = 1e3)
  pvalues_table_contin[i, 2] <- t2p(observed_diff_means, diff_means_distr, alternative = "two-sided")

  observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
  lm1_t_distr <-  replicate(1e3, {
    tmpdata$tr_perm <- permute_within_groups(tmpdata$tr, tmpdata$SITEID)
  lm1_perm <- lm(Outcome ~ Baseline + tr_perm + factor(SITEID), data = tmpdata)
  summary(lm1_perm)[["coefficients"]]["tr_permB", "t value"]
  })
  pvalues_table_contin[i, 3] <- t2p(observed_t, lm1_t_distr, alternative = "two-sided")

  lm_no_tr <- lm(Outcome ~ Baseline + factor(SITEID), data = tmpdata)
  lm_resid <- residuals(lm_no_tr)
  lm_yhat <- fitted(lm_no_tr)
  observed_t <- summary(lm1)[["coefficients"]]["trB", "t value"]
  lm_t_distr <-  replicate(1e3, {
      lm_resid_perm <- permute_within_groups(lm_resid, tmpdata$SITEID)
      tmpdata$response_fl <-  lm_yhat + lm_resid_perm
      lm_perm <- lm(response_fl ~ Baseline + tr + factor(SITEID), data = tmpdata)
      summary(lm_perm)[["coefficients"]]["trB", "t value"]
})

@@
